---
title: "Stat235 Final Project"
author: "Nils Dahlin, Connor Guyette"
date: "10/20/2022"
output: html_document
---
# Final Project Proposal
Nils Dahlin and Connor Guyette

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

pacman::p_load(socviz, tidyverse, rstatix, DescTools, gtExtras, rcompanion, sjPlot)

#ufc_fulldata <- read.csv("data.csv")
#disaster_data <- read.csv("us_disaster_declarations.csv")

drug_data <- read.csv("Drug_Consumption.csv")
```

## Introduction
We are interested in looking at the associations between age/gender/and education and drug usage 
but this data set was also interesting because it measures personality traits such as extroversion, agreeableness, impulsiveness, conscientiousness, and more. We think it would be really interesting to explore the associations between these traits and the use of different drugs and the number of different drugs used

## Data Description
This dataset is a collection of 1884 data points that track different attributes/features of individuals
including age/gender/education level/country/ethnicity/personality trait scores/ and drugs used and when. 

Here is a quick look at the first 5 entries of our dataset.

```{r quick view}
head(drug_data,5)
```

These are the descriptions for the more nondescript features: 

Personality Traits:
NSScore - Neuroticism
EScore - Extraversion
OScore - Openness to Experience
AScore - Agreeableness
CScore - Conscientiousness
Impulsiveness
SS - Sensation Seeing

Drug Use Classifications:
CL0 - Never used
CL1 - Used over a decade ago
CL2 - Used in last decade
CL3 - Used in last year
CL4 - Used in last month
CL5 - Used in last week
CL6 - Used in last day

```{r feature summary}
shortened_df <- select(drug_data,-ID,-Semer,-VSA)
sjPlot::view_df(shortened_df,
 show.frq = T,
 show.prc = T,
 show.na = T,
 show.string.values = T)
```


## Data Summary
We are interested in initially looking at the association between age and certain types of drug use as well as personality traits and drug use. From there we are interested at looking at the associations between these controlling for education level. These are some of our inital data summaries:

### Alcohol Usage vs. Age
This graph shows a barchart of the number of samples for alcohol usage across the 7 classifications further divided by age groups. 
```{r pressure, echo=FALSE}
ggplot(data = drug_data,
       mapping = aes(Alcohol, ..count..)) +
  geom_bar(aes(fill = Age),
           position = "dodge")
```

### Nicotine Usage vs. Age
```{r pressure, echo=FALSE}
ggplot(data = drug_data,
       mapping = aes(Nicotine, ..count..)) +
  geom_bar(aes(fill = Age),
           position = "dodge")
```

### Caffeine Usage vs. Age
```{r pressure, echo=FALSE}
ggplot(data = drug_data,
       mapping = aes(Caff, ..count..)) +
  geom_bar(aes(fill = Age),
           position = "dodge")
```

### Cannabis Usage vs. Age
```{r pressure, echo=FALSE}
ggplot(data = drug_data,
       mapping = aes(Cannabis, ..count..)) +
  geom_bar(aes(fill = Age),
           position = "dodge")
```

### Mushroom/Psylocibin Usage vs. Age
```{r pressure, echo=FALSE}
ggplot(data = drug_data,
       mapping = aes(Mushrooms, ..count..)) +
  geom_bar(aes(fill = Age),
           position = "dodge")
```

### Heroin Usage vs. Age
```{r pressure, echo=FALSE}
ggplot(data = drug_data,
       mapping = aes(Heroin, ..count..)) +
  geom_bar(aes(fill = Age),
           position = "dodge")
```

### Cocaine Usage vs. Age
```{r pressure, echo=FALSE}
ggplot(data = drug_data,
       mapping = aes(Coke, ..count..)) +
  geom_bar(aes(fill = Age),
           position = "dodge")
```

#### Extraversion Score vs Drug Use
```{r}
ggplot(data = drug_data,
       mapping = aes(x=Escore,
                     y=Alcohol,
                     color=Age)) +
  geom_point()
```

